---
title: "Smooth Random Effects Distribution"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{smooth: 01-distribution}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: bibliography.bib
---

```{r global_options, include=FALSE}
knitr::opts_chunk$set(fig.width = 8, fig.height = 4, fig.path='Figs/',
                      echo = FALSE, warning = FALSE, message = FALSE)
```
## Introduction
Longitudinal data where data of individuals are recorded on a
series of occasions are often collected. In fields such as medicine,
pharmacokinetics, microeconomics, and biology among others,
these data allow the researcher to study processes over time. The
linear mixed effects model is a popular model for describing
longitudinal data. These models incorporate fixed effects which
describe relationships between variables and the response that are
valid for the entire population, and random effects drawn from a
population of values which describe relationships associated with
the individual. Here, the variance between individuals is modelled
by the assumed random effects distribution. Usually, the random
effects are assumed to be normally distributed, but this may not
be a realistic assumption. @mixturemixed proposed a
random effects distribution based on a mixture of normal
distributions, leading to a mixture mixed model. A mixture of
normal distributions provides a flexible tool for modelling
non-normal random effects distributions, such as a skewed or
bimodal distribution. In this paper we propose a mixture mixed
model that consists of a large number of normal distributions.
Here, a mixture consisting of ten to twenty components is
specified, in such a way that a smooth random effects distribution
is obtained. A model with too many components may result in
overfitting the data. Therefore, the means of the components are
constrained by placing them a fixed distance apart. A model with
one random effect has a distribution consisting of a series of
normal distributions with means equally spaced on the real line. A
model containing two or more random effects are specified in the
model the means of the mixture distributions are located on a grid
or hyper grid. The distances between the means are
estimated from the data, and we present the additional steps
needed for implementation of the model in an EM algorithm.


## Random Effects Distribution

The random effects distribution in this mixed model is a mixture distribution [@mixturemixed], where the means of the mixture distribution are located on a grid. This means that the distance between  two adjacent means is constant.
As any mixture of normal distributions, the parameters of the components are a probability, a variance-covariance matrix and a mean vector. The variance-covariance matrices are equal across components,  and the means and probabilities are constrained in such a way that they form  a skewed distribution. To obtain this distribution, we impose the following constraints. First, the means of the components are ordered. The distances between two adjacent means are constrained to be equal to a constant $c$, that is, the distances satisfy the relation $u_i - u_{i-1} = c$, for $i = 2,..,K$. 
Second, to ensure monotonicity of the probabilities, we model them through a powerlaw model. Specifically, the probabilities satisfy a truncated power law constraint
 $$ \pi_i = { \theta^i \over \sum \theta ^ i}$$ By specifying the shape of the random effects distribution as skewed to the right or skewed to the left, the probabilities are decreasing or increasing with $i$.
These two constraints are generalized to two dimensions by imposing a grid on the means of the intercept and slope. The joint probability distribution satisfies the independence model because $$\pi_{ij}= \pi_i \pi_j.$$ Third, the variance-covariance matrices are equal across components. A similar distribution is used in [@smr1] where the probabilities are smoothed instead of parametrically modelled as is here the case.

An attractive property of this construction of a random effects distribution is that is contains the normal distribution as a special case. When $u_i - u_{i-1} = 0$ for every $i$, then all components of the mixture are at the same location.

The constraint for the means can be generalized to two dimensions. In this two-dimensional case the means form a grid, where the constraint holds for each dimension. The probabilities satisfy the independence model. The association between the random effects is modelled in the variance covariance matrix and not in the mixture distribution. 



```{r distribution, echo=FALSE}
oldpar <- par()
par(mfrow=c(1, 2))

mus <-1.5*c(0, 1, 2) 

sigma <- 1
xas <- seq(min(mus) - 2.0*sigma, max(mus) + 2.0*sigma, .001)
plot(xas,dnorm(xas, mus[1], sqrt(sigma)), type="l", col="blue", ylab='', 
     xlab=' ', main='Three normal distributions', ylim = c(0, .45))
lines(xas, dnorm(xas, mus[2], sqrt(sigma)), type="l", col="blue")
lines(xas, dnorm(xas, mus[3], sqrt(sigma)), type="l", col="blue")
th <- .25
pow <- c(1:3)
p <- th^pow/sum(th^pow)

plot(xas, p[1]*dnorm(xas, mus[1], sqrt(sigma)), type="l", col="blue", ylab='', 
     xlab=' ',  main='Scaled normal distributions', ylim = c(0, .45))
lines(xas, p[2]*dnorm(xas, mus[2], sqrt(sigma)), type="l", col="blue")
lines(xas, p[3]*dnorm(xas, mus[3], sqrt(sigma)), type="l", col="blue")
```
```{r mixnorm, echo = FALSE}
mixnorm <- function(p, x, mu, sd)
{
  val <- 0
  for (i in 1:length(p))
  {
    val <- val+p[i]*dnorm(x, mu[i], sd)
  }
  val
}
```
The construction of the smooth randomeffects distribution as a mixture of equally spaced normal distributions is illustrated in the figure above. The left panel shows the distribution, and in the right panel the corresponding normal distribution added for comparison.
```{r distribution with normal, echo=FALSE}
oldpar <- par()
par(mfrow=c(1, 2))


plot(xas, mixnorm(p, xas, mus, sigma), type="l", col="red", ylab='',
     xlab=' ', main='Mixture distribution', ylim = c(0, .45))


var <- 0
mu_appr <- sum(p*mus)
for (i in 1:length(p))
{

  var <- var + p[i]*(mus[i]-mu_appr)^2
}
sigma <- sigma + var
plot(xas, mixnorm(p, xas, mus, 1),  type="l", col="red",
     ylab=' ', xlab=' ', main = 'Mixture with normal distribution', ylim = c(0, .45))
lines(xas, dnorm(xas, mu_appr, sqrt(sigma)), type="l", col = "blue")

``` 
The resulting skewed distribution is shown in red, for comparison the normal distribution with the same mean and variance as the mixture distribution is shown in the right panel for comparison.

# References
